---
title: "Loan_Status_randomforest_Project"
author: "Yunwei Chang"
date: "3/9/2021"
output: pdf_document
---

### The accuracy of this random forest model on the test data is 0.9983466 and on the train data is 1.00. 

## Lending Club Data 

```{r}
library(pacman)
p_load(tidyverse, data.table, fs, tictoc, lubridate, beepr,dplyr,skimr, rpart,mdsr, tidymodels,dplyr, forcats)
```

##

```{r}
lending_club_data1 <- fread("lending_club_data_2012_2014.csv")

```

```{r}
head(lending_club_data1)

str(lending_club_data1)
```


```{r}
lending_club_data1 <- lending_club_data1 %>% mutate(year = year(mdy(issue_d)))

lending_club_data1 %>% select(year) %>% 
  distinct(year) %>% 
  arrange(year)
  
lending_club_data_2012_2014 <- lending_club_data1 %>% 
  filter(year %in% c(2012, 2013, 2014))

lending_club_data_2012_2014 %>% select(year) %>% 
  distinct(year) %>% 
  arrange(year)
```


```{r}
dim(lending_club_data_2012_2014)
```



```{r}
lending_club_data_2012_2014$loan_status<-as.factor(lending_club_data_2012_2014$loan_status)
levels(lending_club_data_2012_2014$loan_status)
```


### skim and find variables with lots of NA and remove it later
```{r}
library(skimr)
skim(lending_club_data_2012_2014)
```


### remove columns that have no values and ids
```{r}
lending_club_data_2012_2014 = subset(lending_club_data_2012_2014, select=-c(1,2,19,20,32,39,40,48,50:51,55,56,58,59,60,116:128,131,135:144,146:151))
```


### count how many accounts were created each year
```{r}
lending_club_data_2012_2014%>% group_by(year) %>% 
  count(year)
```

### remove columns with large percentage of NA and are not significant
```{r}
lending_club_data_2012_2014 = subset(lending_club_data_2012_2014, select=-c(27,29,30,48:63,65, 82,85,99))

```


```{r}
lending_club_data_2012_2014_1 <- lending_club_data_2012_2014 %>% filter(loan_status %in% c("Charged Off", "Fully Paid"))

```


### fill in NA with 0
```{r}
lending_club_data_2012_2014_1[is.na(lending_club_data_2012_2014_1)] <- 0

```


### extract a small sample of 500 accounts to test the model.
```{r}
lending_club_data_2012_2014_sample1<-lending_club_data_2012_2014_1[0:1000,]
head(lending_club_data_2012_2014_sample1$loan_status)
```


### only extract first 1000 observations to do data cleaning, splitting, and changing response level to 2
```{r}
lending_club_data_2012_2014_sample1<-lending_club_data_2012_2014_1[0:1000,]
```


```{r}
nlevels(lending_club_data_2012_2014_sample1$loan_status)
status <- c("Charged Off", "Fully Paid", "In Grace Period", "Current", "Default", "Late (31-120 days)", "Late (16-30 days)")
h_gap <- lending_club_data_2012_2014_sample1 %>%
  filter(loan_status %in% status)
nlevels(h_gap$loan_status)
```

### adjust y variable to only 2 levels
```{r}
h_gap_dropped <- h_gap %>% 
  droplevels()
nlevels(h_gap_dropped$loan_status)

h_gap$loan_status<-h_gap$loan_status %>%
  fct_drop() 
levels(h_gap$loan_status)
```



```{r}
library(tidymodels)
set.seed(2)
n <- nrow(h_gap)
h_gap <- h_gap %>%
  initial_split(prop = 0.75)
train1 <- h_gap %>%
  training()
test1 <- h_gap %>%
  testing()
list(train1, test1) %>%
  map_int(nrow)
```


```{r}
library(doMC)
registerDoMC(cores = 8)
```


## Find important terms in numerical variables in these 1000 observations
```{r}
library(Boruta)

set.seed(300)
m_rf_Boruta <- Boruta(loan_status ~ 
 loan_amnt + funded_amnt + funded_amnt_inv + int_rate + installment+ fico_range_low + fico_range_high +  revol_util          
+ total_acc + total_pymnt + total_pymnt_inv+ total_rec_prncp
+ total_rec_int + total_rec_late_fee + recoveries+ collection_recovery_fee+ 
+ last_pymnt_amnt + collections_12_mths_ex_med + acc_now_delinq  + 
tot_coll_amt +  acc_open_past_24mths + bc_util + chargeoff_within_12_mths + delinq_amnt  + mort_acc + mths_since_recent_bc + mths_since_recent_bc_dlq + 
mths_since_recent_inq + mths_since_recent_revol_delinq + num_accts_ever_120_pd +        
  num_bc_tl + num_il_tl + 
num_rev_accts + num_rev_tl_bal_gt_0 + num_sats + 
num_tl_120dpd_2m + num_tl_30dpd + num_tl_90g_dpd_24m + num_tl_op_past_12m
+ pct_tl_nvr_dlq + percent_bc_gt_75 + pub_rec_bankruptcies + tax_liens + 
tot_hi_cred_lim + total_bal_ex_mort + 
total_il_high_credit_limit + hardship_amount, data = train1, doTrace = 2, ntree = 500)


m_rf_Boruta

plot(m_rf_Boruta)

getConfirmedFormula(m_rf_Boruta)

attStats(m_rf_Boruta)

plotImpHistory(m_rf_Boruta)
```

## Find important terms in categorical variables in these observations
```{r}
library(Boruta)

set.seed(300)
m_rf_Boruta2 <- Boruta(loan_status ~ term + grade + sub_grade + emp_title + emp_length + home_ownership + verification_status + issue_d + pymnt_plan + purpose + title + zip_code + addr_state + delinq_2yrs + earliest_cr_line + inq_last_6mths + open_acc + initial_list_status + application_type + hardship_flag + hardship_amount + year , data = train1, doTrace = 2, ntree = 500)


m_rf_Boruta2

plot(m_rf_Boruta2)

getConfirmedFormula(m_rf_Boruta2)

attStats(m_rf_Boruta2)

plotImpHistory(m_rf_Boruta2)
```



## modify response level for the entire training data observations

```{r}
nlevels(lending_club_data_2012_2014_1$loan_status)
status <- c("Charged Off", "Fully Paid", "In Grace Period", "Current", "Default", "Late (31-120 days)", "Late (16-30 days)")
h_gap1 <- lending_club_data_2012_2014_1 %>%
  filter(loan_status %in% status)
nlevels(h_gap$loan_status)
```

### adjust y variable to only 2 levels
```{r}
h_gap_dropped1 <- h_gap1 %>% 
  droplevels()
nlevels(h_gap_dropped1$loan_status)

h_gap1$loan_status<-h_gap1$loan_status %>%
  fct_drop() 
levels(h_gap1$loan_status)
```



```{r}
library(tidymodels)
set.seed(2)
n <- nrow(h_gap1)
h_gap1 <- h_gap1 %>%
  initial_split(prop = 0.75)
train2 <- h_gap1 %>%
  training()
test2 <- h_gap1 %>%
  testing()
list(train2, test2) %>%
  map_int(nrow)
```


### change categorical variables to factors
```{r}
library(tidyverse)
library(magrittr)
cols <- c("term", "grade", "sub_grade", "open_acc", "verification_status", "initial_list_status", "hardship_amount")
train2<-train2 %<>% mutate_at(cols, funs(factor(.)))
```



## Run randomforest with the numerical and categorical variables chosen by Boruta

```{r}
library(randomForest)
library(rpart)

form <- as.formula( loan_status ~ loan_amnt + funded_amnt + funded_amnt_inv + int_rate + 
    installment + total_pymnt + total_pymnt_inv + 
    total_rec_prncp + total_rec_int + total_rec_late_fee + recoveries + 
    collection_recovery_fee + last_pymnt_amnt + last_fico_range_high + last_fico_range_low +       term + grade + sub_grade + verification_status + initial_list_status 
)

loan_tree1 <- decision_tree(mode = "classification") %>%
  set_engine("rpart", control = rpart.control(cp = 0.002)) %>%
  fit(form, data = train2)

mod_forest1 <- rand_forest(
  mode = "classification", 
  mtry = 20, 
  trees = 261
) %>%
  set_engine("randomForest") %>%
  fit(form, data = train2)

```


```{r}

library(yardstick)
pred_sample1 <- train2 %>%
  select(loan_status) %>%
  bind_cols(
    predict(mod_forest1, new_data = train2 , type = "class")
  ) %>%
  rename(loan_tree_matrix = .pred_class)
  
pred_sample1 %>%
  conf_mat(loan_status, loan_tree_matrix)
pred_sample1 %>%
  accuracy(loan_status, loan_tree_matrix)
```

```{r}
library(yardstick)
pred2 <- test2 %>%
  select(loan_status) %>%
  bind_cols(
    predict(mod_forest1, new_data = test2, type = "class")
  ) %>%
  rename(loan_tree_matrix2 = .pred_class)

pred2  %>%
  conf_mat(loan_status, loan_tree_matrix2)
pred2 %>%
  accuracy(loan_status, loan_tree_matrix2)
```



```{r}
fwrite(pred_sample1 ,"prediction of train data")
```


```{r}
fwrite(pred2 ,"prediction of test data")
```

```{r}
pred_sample1
```

```{r}
pred2
```

